Nanophotonics with optical phonons in two dimensions

### 2018-11-07 paper.tex
more theory section

P_i - P_i^{0} = \Omega\frac{\partial P_i}{\partial u_{\kappa,j}}u_{\kappa,j} \equiv Z_{kappa,ij}u_{j},

@@

where $Z_{\kappa,ij}$ is the Born effective charge tensor of ion $\kappa$ and $\Omega$ is the unit cell area. With this relation between polarization and ionic displacements, we have expressed the Hamiltonian as a coupling between the scalar potential and ionic displacements. We consider the response of the monolayer to a scalar potential of the form $\phi = \phi(\mathbf{q},\omega)e^{i\mathbf{q}\cdot\mathbf{r}-i\omega t}$, where $\mathbf{q}$ is a two-dimensional wavevector in the plane of the monolayer. Such a potential corresponds to a longitudinal electric field of the form $\mathbf{E} = i\mathbf{q}\phi(\mathbf{q},\omega)e^{i\mathbf{q}\cdot\mathbf{r}-i\omega t}$. It follows that the induced polarization takes the form $P_i  = \epsilon_0\Pi_{ij}(\mathbf{q},\omega)E_{j,\mathrm{tot}}(\mathbf{q},\omega)$, where $\Pi_{ij}(\mathbf{q},\omega)$ is the polarization-polarization response function of the monolayer and  $E_{j,mathrm{tot}}(\mathbf{q},\omega)$ is taken as the \textit{total} field which is a sum of the applied field and the field created by induced polarization. The polarization-polarization response function is given by

@@

where $m,n$ are states of the phononic Fock space of the monolayer, $P_{i,mn} = \sum\limits_{\kappa,ij}\langle m | u_{\kappa,j} | n \rangle$ are matrix elements of the polarization associated with phonon modes, $E_{m (n)}$ is the energy of state $m (n)$, and $\beta$ is the inverse temperature. Evaluating the contribution of optical phonons in the harmonic approximation to the polarization-polarization response, one finds that the response function is given by: 

@@

To strip the analysis to its bare essentials, we consider optical phonon response with in-plane isotropy in the long-wavelength limit arising from in-plane LO oscillations. A relevant example of a system where these conditions are satisfied is in a hexagonal boron nitride monolayer, considering optical phonons in the so-called upper Reststrahlen band which in bulk spans the frequency range of 1360 to 1610 cm$^{-1}$. In a monolayer geometry with translation invariance and in-plane isotropy, the electromagnetic modes of Maxwell's equations can be composed into transverse magnetic (TM) and transverse electric (TE), where the magnetic or electric field respectively is transverse to the in-plane wavevector of the mode. In practice, it is the TM mode which is associated with highly confined electromagnetic waves that are of use in nanophotonics. TE waves are not supported at the same frequency as transverse magnetic waves, and require exotic conditions to be realized in two-dimensional materials. Given the isotropy, we may suppress indices from the response tensor. We may also consider without further loss of generality a TM mode with wavenumber $q$ along the $x$-direction in the monolayer and magnetic field $H(z)e^{iqx-i\omega t}$ along the $y$-direction of the monolayer. The direction transverse to the monolayer is denoted as $z$.  With these definitions in place, the Maxwell equation satisfied by the magnetic field is \begin{equation}\label{eq:2dmaxwell}
\left(-\frac{\dd^2}{\dd{}z^2}+q^2-\epsilon_{\mathrm{env}}\frac{\omega^2}{c^2} \right)H(z) = 0.
\end{equation}
We consider a solution of the form $H(z) = e^{-\kappa_1 z}$ for $z > 0$ with $\kappa_1 = \sqrt{q^2-\epsilon_1\frac{\omega^2}{c^2}}$ and a solution of the form $H(z) = -ce^{\kappa_2 z}$ for $z < 0$ with $\kappa_2 = \sqrt{q^2-\epsilon_2\frac{\omega^2}{c^2}}$. Continuity of the displacement field across the interface forces $c = \frac{\epsilon_1}{\epsilon_2}$. Meanwhile, the boundary condition on the magnetic field is $\unitvec{z}\times(\mathbf{H}^{(+)}-\mathbf{H}^{(-)}) = \mathbf{K} = -\iu\omega\mathbf{P}_{\mathrm{s}}$ where $\mathbf{K}$ is the surface current density, expressed through the surface polarization density $\mathbf{P}_{\mathrm{s}}$.

The condition for an evanescent electromagnetic wave can be written as $\epsilon_{\mathrm{RPA}}(\omega) = 0$, where $\epsilon_{\mathrm{RPA}}(\omega) = \epsilon_{\mathrm{env}}+\frac{1}{2}q\Pi(\mathbf{q},\omega)$ coincides with the phonon contribution to the longitudinal dielectric function of the monolayer in the random-phase approximation.

>>>>

P_i - P_i^{0} = \Omega\frac{\partial P_i}{\partial u_{\kappa,j}}u_{\kappa,j} \equiv Z_{\kappa,ij}u_{j},

@@

where $Z_{\kappa,ij}$ is the Born effective charge tensor of ion $\kappa$ and $\Omega$ is the unit cell area. With this relation between polarization and ionic displacements, we have expressed the Hamiltonian as a coupling between the scalar potential and ionic displacements. We consider the response of the monolayer to a scalar potential of the form $\phi = \phi(\mathbf{q},\omega)e^{i\mathbf{q}\cdot\mathbf{r}-i\omega t}$, where $\mathbf{q}$ is a two-dimensional wavevector in the plane of the monolayer. Such a potential corresponds to a longitudinal electric field of the form $\mathbf{E} = i\mathbf{q}\phi(\mathbf{q},\omega)e^{i\mathbf{q}\cdot\mathbf{r}-i\omega t}$. It follows that the induced polarization takes the form $P_i  = \epsilon_0\Pi_{ij}(\mathbf{q},\omega)E_{j,\mathrm{tot}}(\mathbf{q},\omega)$, where $\Pi_{ij}(\mathbf{q},\omega)$ is the polarization-polarization response function of the monolayer and  $E_{j,\mathrm{tot}}(\mathbf{q},\omega)$ is taken as the \textit{total} field which is a sum of the applied field and the field created by induced polarization. The polarization-polarization response function is given by

@@

where $m,n$ are states of the phononic Fock space of the monolayer, $P_{i,mn} = \sum\limits_{\kappa,j}\langle m | u_{\kappa,j} | n \rangle$ are matrix elements of the polarization associated with phonon modes, $E_{m (n)}$ is the energy of state $m$ $(n)$, and $\beta$ is the inverse temperature. Evaluating the contribution of optical phonons in the harmonic approximation to the polarization-polarization response, one finds that the response function is given by: 

@@

To strip the analysis to its bare essentials, we consider optical phonon response with in-plane isotropy in the long-wavelength limit arising from in-plane LO oscillations. A relevant example of a system where these conditions are satisfied is in a hexagonal boron nitride monolayer, considering optical phonons in the so-called upper Reststrahlen band which in bulk spans the frequency range of 1360 to 1610 cm$^{-1}$. In a monolayer geometry with translation invariance and in-plane isotropy, the electromagnetic modes of Maxwell's equations can be composed into transverse magnetic (TM) and transverse electric (TE), where the magnetic or electric field respectively is transverse to the in-plane wavevector of the mode. In practice, it is the TM mode which is associated with highly confined electromagnetic waves that are of use in nanophotonics. TE waves are not supported at the same frequency as transverse magnetic waves, and require exotic conditions to be realized in two-dimensional materials. Given the isotropy, we may suppress indices from the response tensor. We may also consider without further loss of generality a TM mode with wavenumber $q$ along the $x$-direction in the monolayer and magnetic field $H(z)e^{iqx-i\omega t}$ along the $y$-direction of the monolayer. The direction transverse to the monolayer is denoted as $z$.  With these definitions in place, the Maxwell equation satisfied by the magnetic field is \begin{equation}\label{eq:2dmaxwell}
\left(-\frac{\dd^2}{\dd{}z^2}+q^2-\epsilon_{\mathrm{env}}\frac{\omega^2}{c^2} \right)H(z) = 0.
\end{equation}
We consider a solution of the form $H(z) = e^{-\kappa_1 z}$ for $z > 0$ with $\kappa_1 = \sqrt{q^2-\epsilon_1\frac{\omega^2}{c^2}}$ and a solution of the form $H(z) = ce^{\kappa_2 z}$ for $z < 0$ with $\kappa_2 = \sqrt{q^2-\epsilon_2\frac{\omega^2}{c^2}}$. The boundary condition on the magnetic field is $H_y^{(+)}-H_y^{(-)} = -K_x = \iu\omega\mathbf{P}_x$ where $\mathbf{K}$ is the surface current density, expressed through the polarization density $\mathbf{P}$ of the monolayer. This condition enforces $c =(1+\frac{\epsilon_0}{\epsilon_2} \kappa_2\Pi)^{-1}$. Continuity of the electric field in the $x$ direction enforces $\frac{\epsilon_2}{\epsilon_1} = -\frac{\kappa_2}{\kappa_1}c$. These two conditions fully specify the TM mode. To simplify the discussion, we consider the so-called electrostatic limit, it in which the wavenumber $q$ of the TM mode is much larger than the free-space wavenumber $\frac{\omega}{c}$. This is well-satisfied in monolayer optical phonons in 2D. With this approximation, the condition for a TM mode corresponding to an optical phonon is
\begin{equation}
\epsilon_{\mathrm{env}} + \frac{1}{2}\epsilon_0 q \Pi(q,\omega) \equiv \epsilon_{\mathrm{RPA}}(\omega) = 0.
\end{equation}
The quantity $\epsilon_{\mathrm{RPA}}(\omega) = 0$ coincides with the phonon contribution to the longitudinal dielectric function of the monolayer in the random-phase approximation.

To proceed, we express the polarization matrix element in Equation (4) in terms of parameters of the monolayer such as the Born effective charges. Considering the longitudinal phonon contribution to the second-quantized ionic displacement, as in [Needs reference!], one finds that the RPA dielectric function is given by:
\begin{equation}
\Pi_{qq}(\mathbf{q},\omega) = \frac{1}{\epsilon_0\hbar\Omega} \frac{2\omega_{\mathbf{q}L}}{\omega^2_{\mathbf{q}TO}-\omega^2}|\hat{q}\cdot\langle 1_{\mathbf{q}L}|\mathbf{P}(\mathbf{q})|0_{\mathbf{q}\sigma}\rangle|^2.
\end{equation}
In this expression, $\omega_L$ and been re-named as $\omega_{TO}$, because in the absence of LO-TO splitting, they are degenerate. Additionally, we have defined eigendisplacements $\boldsymbol{\eta}_{\kappa\mathbf{q}}\equiv \frac{e_{\kappa\mathbf{q},L}}{\sqrt{M_{\kappa}}}$, where $e_{\kappa\mathbf{q},L}$ is the unit-normalized polarization vector of atom $\kappa$ in the unit cell oscillating according to a longitudinal phonon of wavevector $\mathbf{q}$ and $M_{\kappa}$ is the mass of atom $\kappa$.

As is well known, the zeros of the RPA dielectric function correspond to longitudinal electromagnetic waves. Thus, the condition for the zeros of the dielectric function correspond to
\begin{equation}
\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},T} = \frac{e^2}{2\epsilon_0\epsilon_{\mathrm{env}}q}\frac{1}{\Omega}q^2\Big|\sum\limits_{\kappa,j}Z_{\kappa,ij}\eta_{\kappa,j}  \Big|^2
\end{equation}
